---
title: "R Notebook"
output:
  html_document: default
  pdf_document: default
---
Imports 
```{r}
library(ggplot2)
library(tidyverse)
library(magrittr)
library(data.table)
```

Loading the data
```{r}
airlines_data <- read.csv("data/airlines.csv")
airports_data <- read.csv("data/airports.csv")
flights_lax_data <- read.csv("data/flightsLAX.csv")
```



Merge two data tables to include airline names
```{r}
flights <- merge(x=flights_lax_data, y=airlines_data, by.x="AIRLINE", by.y = "IATA_CODE")
flights <- flights %>% 
  rename(
    AIRLINE_CODE = AIRLINE,
    AIRLINE = AIRLINE.y
  )
flights <- as.data.table(flights)
```

Get the flights which have a departure delay and arrival delay
```{r}
departure_delayed_flights <- flights[DEPARTURE_DELAY > 0]
arrival_delayed_flights <- flights[ARRIVAL_DELAY > 0]
```

Top 10 Departure Delay in specific day
```{r}
total_departure_delay_by_day <-  departure_delayed_flights %>% 
  group_by(MONTH,DAY) %>% 
    summarise(sum_departure_delay = sum(DEPARTURE_DELAY, na.rm=TRUE))
# Order the days & get first 10 days
setorder(total_departure_delay_by_day, sum_departure_delay)
first_ten = head(total_departure_delay_by_day,10)

first_ten %>% 
      ggplot(aes(x = paste(DAY,MONTH,sep = '.'), y = sum_departure_delay)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Day",
        y = "Total Departure Delay",
        title = paste(
          "Total 10 Departure Delay by Day"
        )
      )
```

Top 10 Arrival Delay in specific day
```{r}
total_arrival_delay_by_day <-  arrival_delayed_flights %>% 
  group_by(MONTH,DAY) %>% 
    summarise(sum_arrival_delay = sum(ARRIVAL_DELAY, na.rm=TRUE))
# Order the days & get first 10 days
setorder(total_arrival_delay_by_day, sum_arrival_delay)
first_ten = head(total_arrival_delay_by_day,10)

first_ten %>% 
      ggplot(aes(x = paste(DAY,MONTH,sep = '/'), y = sum_arrival_delay)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Day",
        y = "Total Arrival Delay",
        title = paste(
          "Top 10 Total Arrival Delay by Day"
        )
      )
```

Get the average delay by airline company
```{r}
arrival_delayed_flights %>% 
  group_by(AIRLINE, AIRLINE_CODE) %>% 
    summarise(mean_arrival_delay = mean(ARRIVAL_DELAY, na.rm=TRUE)) %>% 
      ggplot(aes(x = AIRLINE_CODE, y = mean_arrival_delay, fill = paste(AIRLINE_CODE,AIRLINE, sep = ' - '))) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airline Company",
        y = "Average arrival delay",
        title = paste(
          "Average arrival delay by companies"
        )
      )
```


```{r}

departure_delayed_flights %>% 
  group_by(AIRLINE_CODE,AIRLINE) %>% 
    summarise(mean_departure_delay = mean(DEPARTURE_DELAY, na.rm=TRUE)) %>% 
      ggplot(aes(x = AIRLINE_CODE, y = mean_departure_delay, fill = paste(AIRLINE_CODE,AIRLINE, sep = ' - '))) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airline Company",
        y = "Average departure delay",
        title = paste(
          "Average departure delay by companies"
        )
      )
```



Departure Delay by Origin Airport
```{r}
departure_delay_by_airport <- departure_delayed_flights[,.(TOTAL_DEPARTURE_DELAY =sum(DEPARTURE_DELAY)),by=ORIGIN_AIRPORT]

departure_delay_by_airport %>%
  ggplot(aes(x = ORIGIN_AIRPORT, y = TOTAL_DEPARTURE_DELAY, fill = ORIGIN_AIRPORT)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airport",
        y = "Total Departure Delay",
        title = paste(
          "Departure Delay by Origin Airport"
        )
      )
```

Average departure delays per flight for each airport
```{r}
departure_delayed_flights %>% 
  group_by(ORIGIN_AIRPORT) %>% 
    summarise(mean_departure_delay = mean(DEPARTURE_DELAY, na.rm=TRUE)) %>% 
      ggplot(aes(x = ORIGIN_AIRPORT, y = mean_departure_delay, fill = ORIGIN_AIRPORT)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airport",
        y = "Average departure delay",
        title = paste(
          "Average departure delay by airports"
        )
      )
```

Arrival Delay by Destination Airport
```{r}
arrival_delay_by_airport <- arrival_delayed_flights[,.(TOTAL_ARRIVAL_DELAY =sum(ARRIVAL_DELAY)),by=DESTINATION_AIRPORT]

arrival_delay_by_airport %>%
  ggplot(aes(x = DESTINATION_AIRPORT, y = TOTAL_ARRIVAL_DELAY, fill = DESTINATION_AIRPORT)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airport",
        y = "Total Arrival Delay",
        title = paste(
          "Arrival Delay by Destination Airport"
        )
      )
```

Average arrival delays per flight for each airport.
```{r}
arrival_delayed_flights %>% 
  group_by(DESTINATION_AIRPORT) %>% 
    summarise(mean_arrival_delay = mean(ARRIVAL_DELAY, na.rm=TRUE)) %>% 
      ggplot(aes(x = DESTINATION_AIRPORT, y = mean_arrival_delay, fill = DESTINATION_AIRPORT)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airport",
        y = "Average Arrival delay",
        title = paste(
          "Average arrival delay by airports"
        )
      )
```


