---
title: "R Notebook"
output:
  html_document: default
  pdf_document: default
  fig_width: 12
---
###### Imports 
```{r}
library(ggplot2)
library(tidyverse)
library(magrittr)
library(data.table)
```

###### Loading the data
```{r}
airlines_data <- read.csv("data/airlines.csv")
airports_data <- read.csv("data/airports.csv")
flights_lax_data <- read.csv("data/flightsLAX.csv")
```



###### Merge two data tables to include airline names
```{r}
flights <- merge(x=flights_lax_data, y=airlines_data, by.x="AIRLINE", by.y = "IATA_CODE")
flights <- merge(x=flights,y=airports_data, by.x = 'ORIGIN_AIRPORT', by.y = 'IATA_CODE')

flights <- flights %>% 
  rename(
    AIRLINE_CODE = AIRLINE,
    AIRLINE = AIRLINE.y,
    ORIGIN_AIRPORT_NAME = AIRPORT
  )

# Merge with airports data to get the destination airport name.
flights <- merge(x=flights,y=airports_data, by.x = 'DESTINATION_AIRPORT', by.y = 'IATA_CODE')


# Rename the destination airport name.
flights <- flights %>% 
  rename(
    DESTINATION_AIRPORT_NAME = AIRPORT
  )

flights <- as.data.table(flights)

# Rename airport and airline to include their abbreviation
flights <- flights[,AIRLINE:=paste(AIRLINE_CODE,AIRLINE,sep = ' - ')]
flights <- flights[,ORIGIN_AIRPORT_NAME:=paste(ORIGIN_AIRPORT,ORIGIN_AIRPORT_NAME,sep = ' - ')]
flights <- flights[,DESTINATION_AIRPORT_NAME:=paste(DESTINATION_AIRPORT,DESTINATION_AIRPORT_NAME,sep = ' - ')]
```

###### Get the flights which have a departure delay and arrival delay
```{r}
departure_delayed_flights <- flights[DEPARTURE_DELAY > 0]
arrival_delayed_flights <- flights[ARRIVAL_DELAY > 0]
```

###### Top 10 Departure Delay in specific day
```{r fig.fullwidth=TRUE}
total_departure_delay_by_day <-  departure_delayed_flights %>% 
  group_by(MONTH,DAY) %>% 
    summarise(sum_departure_delay = sum(DEPARTURE_DELAY, na.rm=TRUE))

# Order the days & get first 10 days
setorder(total_departure_delay_by_day, sum_departure_delay)

# Pick top first 15 days.
o = head(total_departure_delay_by_day,15)

# Plot the ordered graph in decreasing order 
o %>% 
  ggplot(aes(x = reorder(paste(DAY,MONTH,sep = '/'),-sum_departure_delay), y = sum_departure_delay)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  labs(
    x = "Day",
    y = "Total Departure Delay",
    title = paste(
      "Top 15 Total Departure Delay by Day (Whole Country)"
    )
  )
```
     
> As we can see 31st of December has the most departure delays maybe due to New Years Eve. 

###### Top 10 Arrival Delay in specific day
```{r}
total_arrival_delay_by_day <-  arrival_delayed_flights %>% 
  group_by(MONTH,DAY) %>% 
    summarise(sum_arrival_delay = sum(ARRIVAL_DELAY, na.rm=TRUE))
# Order the days & get first 10 days
setorder(total_arrival_delay_by_day, sum_arrival_delay)

# Pick the first 15 case.
o = head(total_arrival_delay_by_day, 15)

o %>% 
  ggplot(aes(x = reorder(paste(DAY,MONTH,sep = '/'),-sum_arrival_delay), y = sum_arrival_delay)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  labs(
    x = "Day",
    y = "Total Arrival Delay",
    title = paste(
      "Top 15 Total Arrival Delay by Day (Whole Country)"
    )
  )
```

###### Get the average departure delay by airline
```{r}
o <- departure_delayed_flights %>% 
  group_by(AIRLINE_CODE,AIRLINE) %>% 
    summarise(mean_departure_delay = mean(DEPARTURE_DELAY, na.rm=TRUE))

o %>% 
  ggplot(aes(x = reorder(AIRLINE_CODE, -mean_departure_delay), y = mean_departure_delay, fill = AIRLINE)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  labs(
    x = "Airline Company",
    y = "Average departure delay",
    title = paste(
      "Average departure delay by airline"
    )
    )
```
     
  > American Eagle Airlines has the most average departure delay. 

###### Get the average arrival delay by airline company
```{r}
arrival_delayed_flights %>% 
  group_by(AIRLINE, AIRLINE_CODE) %>% 
    summarise(mean_arrival_delay = mean(ARRIVAL_DELAY, na.rm=TRUE)) %>% 
      ggplot(aes(x = reorder(AIRLINE_CODE, -mean_arrival_delay), y = mean_arrival_delay, fill = AIRLINE)) +
      geom_bar(stat = "identity") +
      theme_classic() +
      labs(
        x = "Airline Company",
        y = "Average arrival delay",
        title = paste(
          "Average arrival delay by companies"
        )
      )
```

 > American Eagle Airlines has the most average arrival delay. 

###### Departure Delay by Origin Airport
```{r}
o <- departure_delayed_flights %>%
  group_by(ORIGIN_AIRPORT, ORIGIN_AIRPORT_NAME) %>%
    summarise(mean_departure_delay = sum(DEPARTURE_DELAY, na.rm=TRUE)) 
setorder(o,-mean_departure_delay)

# Pick the top entries
o <- head(o,10)

o %>%
  ggplot(aes(x = reorder(ORIGIN_AIRPORT, -mean_departure_delay), y = mean_departure_delay, fill = ORIGIN_AIRPORT_NAME)) +
    geom_histogram(stat = "identity") +
    theme_classic() +
    labs(
      x = "Airport",
      y = "Total departure delay",
      title = paste(
        "Top 10 departure total delays by airports"
      )
    )
```

###### Average departure delays per flight for each airport
```{r}
o <- departure_delayed_flights %>%
  group_by(ORIGIN_AIRPORT, ORIGIN_AIRPORT_NAME) %>%
    summarise(mean_departure_delay = mean(DEPARTURE_DELAY, na.rm=TRUE)) 
setorder(o,-mean_departure_delay)
o <- head(o,10)

o %>%
  ggplot(aes(x = reorder(ORIGIN_AIRPORT,-mean_departure_delay), y = mean_departure_delay, fill = ORIGIN_AIRPORT_NAME)) +
    geom_histogram(stat = "identity") +
    theme_classic() +
    labs(
      x = "Airport",
      y = "Average departure delay",
      title = paste(
        "Average departure delay by airports, top 10"
      )
    )
```
    
  > 
    - As we see MTJ - Montrose Regional Airport has the most average departure delay
    - Now we see that LAX is not even in the list because it is a good airport.  

###### Arrival Delay by Destination Airport
```{r}
o <- arrival_delayed_flights %>%
  group_by(DESTINATION_AIRPORT, DESTINATION_AIRPORT_NAME) %>%
    summarise(mean_arrival_delay = sum(ARRIVAL_DELAY, na.rm=TRUE)) 
setorder(o,-mean_arrival_delay)

# Pick the top entries
o <- head(o,10)

o %>%
  ggplot(aes(x = reorder(DESTINATION_AIRPORT, -mean_arrival_delay), y = mean_arrival_delay, fill = DESTINATION_AIRPORT_NAME)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    labs(
      x = "Airport",
      y = "Total arrival delay",
      title = paste(
        "Top 10 arrival total delays by airports"
      )
    )
```


###### Average arrival delays per flight for each airport.
```{r}
o <- arrival_delayed_flights %>%
  group_by(DESTINATION_AIRPORT, DESTINATION_AIRPORT_NAME) %>%
    summarise(mean_arrival_delay = mean(ARRIVAL_DELAY, na.rm=TRUE)) 
setorder(o,-mean_arrival_delay)
o <- head(o,10)

o %>%
  ggplot(aes(x = reorder(DESTINATION_AIRPORT,-mean_arrival_delay), y = mean_arrival_delay, fill = DESTINATION_AIRPORT_NAME)) +
    geom_histogram(stat = "identity") +
    theme_classic() +
    labs(
      x = "Airport",
      y = "Average arrival delay",
      title = paste(
        "Average arrival delay by airports, top 10"
      )
    )
```
   
   > Cleveland Hopkins International Airport has the most average arrival delay.

